---
title: "Own-source revenue for pension stress tests"
author: "Don Boyd"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook:
    df_print: paged
    fig_height: 6
    fig_width: 8
    toc: yes
  html_document:
    toc: yes
    df_print: paged
editor_options:
  chunk_output_type: inline
---


```{r setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=FALSE)
options(width = 150)
```


```{r globals}


```


```{r libraries, include=FALSE}
source(here::here("include", "libraries.r"))
devtools::session_info()
```

<!--
Note:

-->


# NJ forecast - introduction

Goal is to forecast NJ own-source revenue (osr) and components using a method that:

*   is compatible with a state's own short-term forecasts,
*   stabilizes in the longer term to grow at the rate of state gdp, and
*   can be applied to other states with a minimal amount of digging for new information


Key forecast elements:

*   NJ own-source revenue and major components
    +   using Census definitions
    +   state fiscal years (periods ending in June)
    +   nominal
    +   $ millions
    +   2021 through 2030
    
*   National economic forecast - uses one of the relatively current forecasts provided by Pew:
    +   CBO July 2020 (baseline)
    +   Moody's July 2020, p50
    +   Moody's July 2020, p90 (S3 - severe stress scenario)
    +   (We also have available a Moody's December 2019 baseline forecast, not used)
    +   Federal Reserve alternative severe forecast
    
*   NJ economic forecast:
    +   available state GDP forecasts are the Moody's December 2019 (which is not used below) and Moody's July 2020 p50 (baseline) forecasts
    +   the only pair of US-NJ GDP forecasts we have that are directly comparable are the Moody's July 2020 p50 national and state forecasts
    +   I compute the difference between these national and state GDP forecasts and assume that the difference between national and NJ growth for other national forecasts is the same as this
    

Overall approach for OSR and components:

*   2019 and earlier - Census
*   2020 - tax growth rates from Census quarterly data; nontax based on state GDP estimate
*   2021 uses growth rates from a NJ-governmental forecast:
    +   currently uses the governor's August 2020 estimate
    +   note that governor raised his estimate in Nov 2020 by [$398 million](https://www.nj.gov/treasury/news/2020/11062020.shtml); this is not yet reflected but can be
    +   also note that OLS estimate is about $1.3 billion higher than gov August, could be used as an alternative forecast for 2021
*   2022-2024 taxes move in equal annual increments, as % of GDP, from their 2021 % of GDP to their presumed long-term % of GDP (the presumed long-term % of GDP is the average observed for the 3 years ending in 2019)
    +   thus, sources that declined most sharply will bounce back most sharply
*   2025-2030 - all osr components are a constant % of GDP - their presumed long-term average (thus, all osr components grow at the same rate as state GDP)


Selected issues and considerations:

*   It is a forecast under a single scenario - does not consider possible impact of stock market declines, for example; these are possible extensions
*   Uses Census definitions of OSR and thus is different from what policymakers in NJ are familiar with
*   Assumes all revenue sources move from their 2021 values to longer term values, as % of GDP; historically, other tax revenue has tended to decline as % of GDP until policymakers explicitly raise it
*   Because the 2021 OSR estimate is based on a fixed governmental forecast, and is independent of the economic forecast, it may seem odd in relation to a severe-downturn economic forecast for FY 2021 - in particular, OSR as a percent of GDP is unusually high in 2021 when using the Moody's p90 forecast (because gdp is so low in that forecast)
*   Because OSR as % of GDP was rising in years leading up to recession, it reverts to a high level. This reflects recent tax-increase choices, I think.


# NJ forecast - results


```{r parameters, include=FALSE}
nontax_growth2021 <- 0  # must assume as we don't have a forecast, NJ growth rate
pctgdp_years <- 3 # number of years we use to determine osr target pct of gdp
recover_years <- 4  # number of years from trough to return to target % of gd p

gov_forecast <- "govaugxleg"  # govbmxleg, govaugxleg, olssepxleg

osrvars <- c("iit", "gst", "cit", "othertax", "nontax") # define the desired order

keyvars <- c("cit_census", "gst_census", "iit_census", "othertax_census",
           "nontax_census", "osr_census", "gdp_beaspliced")

```


```{r getdata, include=FALSE}
combo <- readRDS(here::here("data", "combo.rds"))

combotab <- combo %>%
  group_by(fullname) %>%
  summarise(n=n(),
            n_notNA=sum(!is.na(value)),
            first_year=min(year),
            last_year=max(year),
            nyears=last_year - first_year + 1,
            nstates=length(unique(stabbr)),
            nus=sum(stabbr=="US", na.rm = TRUE),
            nnj=sum(stabbr=="NJ", na.rm = TRUE),
            minval=min(value, na.rm=TRUE),
            maxval=max(value, na.rm=TRUE),
            .groups="drop") %>%
  select(fullname, n, n_notNA, nyears, everything())
combotab

```


```{r gov_forecasts, include=FALSE}
# we need 2021 growth rates for each source
# note that we just need iit, gst, cit, othertax, and nontax growth rates
# we'll compute total growth rates and osr growth rates

# govaugxleg and olssepxleg are possible forecasts
taxgrowth <- combo %>%
  filter(str_detect(fullname, "xleg")) %>%
  group_by(src, name) %>%
  arrange(year) %>%
  mutate(pch=value / value[match(year - 1, year)] * 100 - 100) %>%
  ungroup %>%
  filter(year == 2021) %>%
  select(year, name, pch, src) %>%
  pivot_wider(values_from = pch)

# add the nontax growth rate
taxgrowth <- taxgrowth %>%
  mutate(nontax=nontax_growth2021)

taxgrowth

taxgrowth_long <- taxgrowth %>%
  pivot_longer(-c(year, src), values_to = "pch")

```



```{r osr_LR_target, include=FALSE}

# Compute long-run target revenue as % of gsp

pctgdp <- combo %>%
  filter(fullname %in% keyvars, stabbr=="NJ", year >= 1964, year <= 2019) %>% # so we always have gdp
  select(year, name, value) %>%
  group_by(year) %>%
  mutate(gdp=value[name=="gdp"], 
         pctgdp=value / gdp * 100) %>%
  filter(name != "gdp") %>%
  group_by(name) %>%
  arrange(year) %>%
  mutate(mapct=ma(pctgdp, n=pctgdp_years)) %>%
  select(year, name, mapct) %>%
  pivot_wider(values_from = mapct)

# pctgdp %>%
#   filter(year >= 1980) %>%
#   pivot_longer(-year) %>%
#   filter(!name %in% c("osr")) %>%
#   ggplot(aes(year, value, colour=name)) +
#   geom_line() +
#   geom_point()

targets <- pctgdp %>%
  filter(year==max(year)) %>%
  pivot_longer(-year, values_to = "targetpct")
targets

```



```{r forecast, include=FALSE}
# Put it all together
# history to 2020
# govt growth rates to 2021
# phase toward the long run over x years
# then implement the long run

# we need our spliced national and state gdp forecasts
usgdp <- readRDS(here::here("data", "usgdp_spliced.rds"))
njgdp <- readRDS(here::here("data", "njgdp_spliced.rds"))

(start_year <- max(min(usgdp$year), min(njgdp$year)))

econdata <- bind_rows(usgdp %>%
                        mutate(name="gdp", stabbr="US") %>%
                        select(year, name, stabbr, value, src),
                      njgdp %>%
                        select(year, name, stabbr, value, src)) %>%
  filter(year >= start_year)
count(econdata, name)
count(econdata, stabbr)
count(econdata, src)
econw <- econdata %>%
  pivot_wider(names_from = stabbr)

# now we need tax data, which will be the same
osrtax <- combo %>%
  filter(src=="census", stabbr=="NJ", year >= min(econw$year), name %in% osrvars) %>%
  select(year, name, value) %>%
  mutate(name=factor(name, levels=osrvars)) %>%
  arrange(year, name) %>%
  pivot_wider() %>%
  replace_na(list(cit = 0))

fcdata <- econw %>%
  filter(name=="gdp") %>%
  select(year, src, gdpUS=US, gdpNJ=NJ) %>%
  left_join(osrtax, by="year")
summary(fcdata)

# compute 2021 based on government forecast
d2021 <- fcdata %>%
  pivot_longer(-c(year, src, gdpUS, gdpNJ)) %>%
  left_join(taxgrowth_long %>% 
              filter(src==gov_forecast) %>% 
              select(year, name, pch), 
            by = c("year", "name")) %>%
  group_by(src, name) %>%
  mutate(value=ifelse(year==2021, value[year==2020] * (1 + pch / 100), value),
         pctgdp=value / gdpNJ * 100)
  
# now compute % gdp and gap vs target
# define # of years it will take to return to target
fcast <- d2021 %>%
  select(-pch) %>%
  left_join(targets %>% select(name, targetpct),
            by="name") %>%
  mutate(gap=ifelse(year >= 2015, pctgdp - targetpct, NA_real_),
         years_from_start=case_when(year <= 2021 ~ 0,
                              year  %in% 2022:(2021 + recover_years) ~ year - 2021,
                              TRUE ~ 0),
         initial_gap = gap[year==2021],
         initial_pct = pctgdp[year==2021],
         pctgdp_new=case_when(year <= 2021 ~ NA_real_,
                              year %in% 2022:(2021 + recover_years) ~ 
                                initial_pct - initial_gap * years_from_start / recover_years,
                              year > (2021 + recover_years) ~ targetpct,
                              TRUE ~ -999),
         value=ifelse(year >= 2022,
                      gdpNJ * pctgdp_new / 100,
                      value)) %>%
  ungroup


# fcast %>%
#   filter(name=="gst", year >= 2010, src!="cbo_precovid") %>%
#   ggplot(aes(year, value, colour=src)) +
#   geom_line() +
#   geom_point()


fc_wide <- fcast %>%
  select(year, src, gdpUS, gdpNJ, name, value) %>%
  arrange(src, year, name) %>%
  pivot_wider() %>%
  mutate(tottax=iit + gst + cit + othertax,
         osr=tottax + nontax)
  
# fcast %>%
#   filter(src!="cbo_precovid", year >= 2015) %>%
#   ggplot(aes(year, osr, colour=src)) +
#   geom_line() +
#   geom_point()

# fc_wide %>%
#   filter(src!="cbo_precovid", year >= 2015) %>%
#   ggplot(aes(year, osr / gdpNJ * 100, colour=src)) +
#   geom_line() +
#   geom_point()

# fcast %>%
#   group_by(src) %>%
#   mutate(osrpct=osr / osr[match(year - 1, year)] * 100 - 100) %>%
#   filter(src!="cbo_precovid", year >= 2015) %>%
#   ggplot(aes(year, osrpct, colour=src)) +
#   geom_line() +
#   geom_point() +
#   geom_hline(yintercept = 0) +
#   scale_x_continuous(breaks=seq(2010, 2040, 2)) +
#   scale_y_continuous(breaks=seq(-20, 20, 2))


```


## Tables

```{r prep, include=FALSE}
vars <- c("iit", "gst", "cit", "othertax", "tottax", "nontax", "osr")

```


### CBO July 2020 forecast

#### Dollar amounts
```{r include=TRUE}
data <- fc_wide %>%
  filter(src=="cbo_jul2020", year >= 2019) %>%
  select(year, any_of(vars))

tab <- data %>%
  gt() %>%  
  tab_header(
      title = "New Jersey own-source revenue, $ millions",
      subtitle = "National forecast: CBO July 2020 baseline"
    ) %>%
    cols_label(
      year=html("State fiscal year<br>ending in:")
    ) %>%
    fmt_currency(
      columns = vars,
      rows=1,
      decimals = 0,
      scale_by = 1,
      suffixing = FALSE
    ) %>%
    fmt_number(
      columns = vars,
      rows=2:nrow(data),
      decimals = 0,
      scale_by = 1,
      suffixing = FALSE
    ) %>%
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
    tab_source_note("OSR categories use Census-definitions")

tab

```

### % Change
```{r include=TRUE}
data <- fc_wide %>%
  filter(src=="cbo_jul2020", year >= 2018) %>%
  select(year, any_of(vars)) %>%
  arrange(year) %>%
  mutate(across(any_of(vars), ~ . / lag(.) * 100 - 100)) %>%
  filter(year >= 2019)

tab <- data %>%
  gt() %>%  
  tab_header(
      title = "New Jersey own-source revenue, % change",
      subtitle = "National forecast: CBO July 2020 baseline"
    ) %>%
    cols_label(
      year=html("State fiscal year<br>ending in:")
    ) %>%
    fmt_percent(
      columns = vars,
      # rows=1:nrow(data),
      decimals = 1,
      scale_values = FALSE
    ) %>%
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
    tab_source_note("OSR categories use Census-definitions")

tab
```



## Graphs

Note: In the graphs that follow, all series have the same (actual) history but only one of the lines will be visible.

### Forecast comparison - levels
```{r include=TRUE}
fc_wide %>%
  filter(year >= 2015, src!="cbo_precovid") %>%
  ggplot(aes(year, osr, colour=src)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(name="$ millions", breaks=seq(20e3, 80e3, 5e3), labels=comma) +
  scale_x_continuous(name="State fiscal year", breaks=seq(2010, 2040, 2)) +
  ggtitle("New Jersey own-source revenue, $ millions",
          subtitle="Alternative national economic forecasts") +
  theme_bw()

```


### Forecast comparison - % change
```{r include=TRUE}
fc_wide %>%
  group_by(src) %>%
  mutate(pch=osr / osr[match(year - 1, year)] * 100 - 100) %>%
  filter(year >= 2015, src!="cbo_precovid") %>%
  ggplot(aes(year, pch, colour=src)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0) +
  scale_y_continuous(name="% change", breaks=seq(-10, 20, 1)) +
  scale_x_continuous(name="State fiscal year", breaks=seq(2010, 2040, 2)) +
  ggtitle("New Jersey own-source revenue, % change from previous year",
          subtitle="Alternative national economic forecasts") +
  theme_bw()

```


### Forecast comparison - % of GDP
```{r include=TRUE}
fc_wide %>%
  group_by(src) %>%
  filter(year >= 2010, src!="cbo_precovid") %>%
  mutate(pctgdp=osr / gdpNJ * 100) %>%
  ggplot(aes(year, pctgdp, colour=src)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(name="% of state GDP", breaks=seq(1, 20, 0.2)) +
  scale_x_continuous(name="State fiscal year", breaks=seq(2010, 2040, 2)) +
  ggtitle("New Jersey own-source revenue, % of state GDP",
          subtitle="Alternative national economic forecasts") +
  theme_bw()

```


