---
title: "Own-source revenue for pension stress tests"
author: "Don Boyd"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook:
    df_print: paged
    number_sections: true
    fig_height: 6
    fig_width: 8
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: true
      smooth_scroll: true
editor_options:
  chunk_output_type: inline
---


```{r setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include=FALSE, fig.width=10, fig.height=6)
options(width = 150)
```


```{r libraries, include=FALSE}
source(here::here("include", "libraries.r"))
devtools::session_info()
```


```{r globals}

  # html_document:
  #   df_print: paged
  #   fig_height: 6
  #   fig_width: 8
  #   toc: yes
  #   toc_depth: 5
  #   toc_float:
  #     collapsed: true
  #     smooth_scroll: false      


  # html_document:
  #   toc: yes
  #   toc_depth: 5
  #   toc_float:
  #     collapsed: false
  #     smooth_scroll: false
  #   df_print: paged
```



<!--
Note:

-->


**See explanations AFTER the graphs and tables below, in the *Goals and methods section*.**


# Preliminaries

```{r functions}
pch <- function(df){
  df2 <- df %>%
    mutate(pch=value / value[match(year - 1, year)] * 100 - 100)
  df2$pch
}

stname <- function(stabbr){
  state.name[state.abb == stabbr]
}

```


```{r get_data}
# get data
usgdp <- readRDS(here::here("data", "usgdp.rds"))  # $ millions
stgdp <- readRDS(here::here("data", "stgdp.rds"))  # $ millions

sp500tr3 <- readRDS(here::here("data", "sp500tr.rds"))
cg <- readRDS(here::here("data", "uscapgains.rds"))
cgmodel <- readRDS(here::here("data", "cgmodel.rds"))
cgfcast <- readRDS(here::here("data", "cgfcast.rds"))

osr <- readRDS(here::here("data", "osr_census_updated.rds")) %>% # ends 2020
  mutate(value=value)  %>% 
  arrange(stabbr, name, year) %>%
  group_by(stabbr, name) %>%
  mutate(pch=value / value[match(year - 1, year)] * 100 - 100) %>%
  ungroup
summary(osr)  # $ millions


osr_pctgdp <- readRDS(here::here("data", "osr_pctgdp.rds"))

targets <- osr_pctgdp %>%
  pivot_longer(cols=starts_with("p"), values_to = "longrun_targetpct", names_to = "ma_years") %>%
  mutate(ma_years=str_sub(ma_years, 2, -1) %>% as.integer)
# targets

pewosr <- readRDS(here::here("data", "pew_osrforecasts.rds"))

```



```{r get_assumptions, include=FALSE}
fn <- "assumptions.xlsx"

assume <- read_excel(here::here("data", fn), sheet="osr_forecasts") %>%
  filter(!is.na(stabbr)) %>%
  # convert to percentage
  mutate(across(c(iit, cit, gst, othertax, nontax), ~ . * 100))
assume

```


## Scenarios and assumptions

```{r define_scenarios}
# define scenarios as combinations of:

scenarios <- tribble(
  ~stabbr, ~scenario, ~econfc, ~equityfc, ~govtfc,

  "MA",  "precovid", "cbo_precovid", "pew_covidbase", "consensus_jan2020",
  "MA",  "baseline", "cbo_feb2021", "pew_covidbase", "gov_oct2020",
  "MA",  "downside", "fedalt", "pew_fedalt", "gov_oct2020", 
  
  "MN",  "precovid", "cbo_precovid", "pew_covidbase", "gov_feb2020",
  "MN",  "baseline", "cbo_feb2021", "pew_covidbase", "gov_nov2020",
  "MN",  "downside", "fedalt", "pew_fedalt", "gov_nov2020",
  
  "NJ",  "precovid", "cbo_precovid", "pew_covidbase", "gov_bmxleg",  
  "NJ",  "baseline", "cbo_feb2021", "pew_covidbase",  "ols_sep2020",
  "NJ",  "downside", "fedalt", "pew_fedalt", "gov_aug2020",
)

# scenarios <- tribble(
#   ~stabbr, ~scenario, ~econfc, ~equityfc, ~govtfc,
# 
#   "MA",  "precovid", "cbo_precovid", "pew_covidbase", "consensus_jan2020",
#   "MA",  "baseline", "moodys_dec2020", "pew_covidbase", "gov_oct2020",
#   "MA",  "downside", "fedalt", "pew_fedalt", "gov_oct2020", 
#   
#   "MN",  "precovid", "cbo_precovid", "pew_covidbase", "gov_feb2020",
#   "MN",  "baseline", "moodys_dec2020", "pew_covidbase", "gov_nov2020",
#   "MN",  "downside", "fedalt", "pew_fedalt", "gov_nov2020",
#   
#   "NJ",  "precovid", "cbo_precovid", "pew_covidbase", "gov_bmxleg",  
#   "NJ",  "baseline", "moodys_dec2020", "pew_covidbase",  "ols_sep2020",
#   "NJ",  "downside", "fedalt", "pew_fedalt", "gov_aug2020",
# )


```


```{r define_forecast_parameters}

trough_year <- 2021

model_forecast_years <- 2022:2025   # period for model forecasts
gdpgap_recovery_years <- 2022:2025  # period for gap-closing forecasts, may not be same as model_forecast_years

```


Scenarios examined:
```{r include=TRUE}
scenarios

```


Technical parameters:

```{r include=TRUE}
cat("Trough year:")
trough_year

cat("\nModel forecast years:\n")
model_forecast_years

cat("\nYears over which OSR recovers as % of GDP in the gap-closing approach:\n")
gdpgap_recovery_years

```



## National economic forecasts

The plots below show:

*   All available nominal GDP forecasts
*   Pre-Covid forecasts: CBO early 2020 (January?) and Moody's December 2019
*   Moody's baseline forecasts and how they have changed over time (December 2019, July 2020, December 2020)
*   Downside Covid forecasts (Moody's July 2020 p90 severe stress scenario, and the Federal reserve alternative fedalt forecast)

The forecasts are on a July-June state fiscal year basis. (Each year ends in June. For example, the 2021 forecast is for the 12 months ending in June 2021.)


```{r}
plot_fc <- function(fcorder, titlepre){
  usgdp %>%
    select(year, all_of(fcorder)) %>%
    filter(year >= 2010) %>%
    pivot_longer(-year, names_to = "forecast") %>%
    mutate(forecast=factor(forecast, level=fcorder)) %>%
    arrange(forecast) %>%
    ggplot(aes(year, value / 1e3, colour=forecast)) +
    geom_line(size=1) +
    geom_point() +
    scale_y_continuous(name="$ billions", breaks=seq(10e3, 50e3, 2.5e3), labels=scales::comma) +
    scale_x_continuous(name="State fiscal year", breaks=seq(2010, 2040, 2)) +
    ggtitle(paste0(titlepre, " national forecasts of nominal GDP, state fiscal year basis")) +
    theme_bw()
}

```


### All available national forecasts
```{r include=TRUE, fig.width=10, fig.height=6}

fcorder <- c("cbo_precovid", "moodys_dec2019", "cbo_jul2020", "moodys_jul2020p50",
             "moodys_dec2020", "cbo_feb2021",
             "moodys_jul2020p90", "fedalt")
plot_fc(fcorder, "All")

```


### Pre-Covid national forecasts
```{r include=TRUE, fig.width=10, fig.height=6}

fcorder <- c("cbo_precovid", "moodys_dec2019")
plot_fc(fcorder, "Pre-Covid") +
  scale_colour_manual(values=c("blue", "darkgreen"))

```


### Moody's baseline national forecasts, various release dates
```{r include=TRUE, fig.width=10, fig.height=6}

fcorder <- c("moodys_dec2019", "moodys_jul2020p50", "moodys_dec2020")
plot_fc(fcorder, "Moody's baseline")

```

### Most-recent Moody's and CBO national forecasts
```{r include=TRUE, fig.width=10, fig.height=6}

fcorder <- c("cbo_feb2021", "moodys_dec2020")
plot_fc(fcorder, "Most recent baseline") +
  scale_colour_manual(values=c("blue", "darkgreen"))

```


### Downside Covid national forecasts
```{r include=TRUE, fig.width=10, fig.height=6}

fcorder <- c("moodys_jul2020p90", "fedalt")
plot_fc(fcorder, "Downside Covid") +
  scale_colour_manual(values=c("blue", "darkgreen"))

```


## National capital gains forecasts

Capital gains are important to state income taxes, as will be apparent when we look at individual states.

We have a small model that estimates and forecasts national capital gains based upon national GDP, the stock market (S&P 500 total returns), and, in effect, capital gains' own history. (A regression with ARIMA errors model.)

The model is an ma1 model, which means that the estimate of capital gains for a given year reflects:
- gdp: GDP for the year
- sp500tr: stock market value for the year
- the error in the estimate for the previous year; because this is positive in the fit statistics shown below, it means there is autocorreation - for example, an unusually large value for capital gains in one year is likely to be followed by an unusually large value in the next


Here are fit statistics from the model:


```{r include=TRUE}
report(cgmodel)
```

Here is a plot of how well/poorly the model does over history. It captures broad trends but is not very accurate, as is generally true of capital gains models: capital gains are highly volatile and difficult to predict.

```{r cgfit, include=TRUE, fig.width=10, fig.height=6}

cgmodel %>% 
  augment() %>%
  select(year, actual=cg, fitted=.fitted) %>%
  pivot_longer(-year, names_to = "type") %>%
  ggplot(aes(year, value, colour=type)) +
  geom_line(size=1) +
  geom_point() +
  scale_colour_manual(values=c("blue", "darkgreen")) +
  scale_x_continuous(name="Calendar year", breaks=seq(1990, 2020, 2)) +
  scale_y_continuous(name="$ billions", labels = scales::comma) +
  ggtitle("Model fit for national capital gains model") +
  theme_bw()

```

We used this model to forecast capital gains under different economic forecasts, using the baseline and fedalt equity return assumptions. Here are the forecasts under selected economic forecast and equity-return assumption combinations.

The combination of the fedalt economic scenario plus the fedalt equity-return assumptions results in large capital gains reductions and will have large impacts on state personal income tax forecasts.


```{r cgfc, include=TRUE, fig.width=10, fig.height=6}

cgfcast %>%
  filter(year >= 2010) %>%
  filter((gdpname=="cbo_precovid" & eqname=="pew_covidbase") |
         (gdpname=="cbo_jul2020" & eqname=="pew_covidbase") |
         (gdpname=="moodys_dec2020" & eqname=="pew_covidbase") |
         (gdpname=="fedalt" & eqname=="pew_covidbase") |
         (gdpname=="fedalt" & eqname=="pew_fedalt")) %>%
  mutate(grp=paste0(gdpname, "-", eqname)) %>%
  ggplot(aes(year, cgfcast, colour=grp)) +
  geom_line(size=1) +
  geom_point() +
  scale_x_continuous(name=NULL, breaks=seq(1990, 2040, 2)) +
  scale_y_continuous(name="$ billions", labels = scales::comma) +
  ggtitle("Model forecasts of capital gains",
          subtitle="Selected economic-equity forecast combinations") +
  theme_bw()


```


# State Own-Source Revenue analysis and forecasts

```{r state_data_prep}
# prep data for ALL states, then draw as needed for individual states

# state and US gdp, history and forecast, all economic forecasts, 1948-2030
stusgdp <- bind_rows(usgdp %>% mutate(stabbr="US"),
                     stgdp) %>%
  pivot_longer(-c(year, stabbr), names_to = "forecast") %>%
  group_by(forecast, stabbr) %>%
  mutate(pch=value / value[match(year - 1, year)] * 100 -100) %>%
  ungroup

# same, but wide (no rows for US), 1965-2030
stgdp_uswide <- stusgdp %>%
  group_by(forecast, year) %>%
  mutate(usgdp=value[stabbr=="US"],
         usgdp_pch=pch[stabbr=="US"]) %>%
  ungroup %>%
  filter(stabbr != "US", year >= 1965) %>%
  rename(stgdp=value, stgdp_pch=pch)

# cg history and forecast, economic and equity forecast combinations, 1995-2030
capgains <- cgfcast %>%
  select(year, gdpname, eqname, capgains=cgfcast) %>%
  group_by(gdpname, eqname) %>%
  mutate(capgains_lag=capgains[match(year - 1, year)],
         capgains_pch=capgains / capgains_lag * 100 - 100) %>%
  ungroup
count(capgains, gdpname, eqname)

# state gdp and osr pch, 1965-2020
gdposr_pch <- stgdp %>%
  select(stabbr, year, gdp=moodys_dec2020) %>%
  inner_join(osr %>% 
               filter(name=="osr") %>% 
               select(stabbr, year, osr=value),
             by = c("stabbr", "year")) %>%
  pivot_longer(cols=-c(year, stabbr)) %>%
  group_by(stabbr, name) %>%
  mutate(pch=value / value[match(year - 1, year)] * 100 - 100) %>%
  ungroup %>%
  filter(year > min(year))

# State GDP, capital gains, and income tax, 1995-2020, based on moodys_dec2020
# on expectation that it will give the best estimate of FY 2020 gdp
iit_gdpcg <- stgdp %>%
  select(year, stabbr, gdp=moodys_dec2020) %>%
  inner_join(osr %>% 
               filter(name=="iit") %>% 
               select(stabbr, year, iit=value),
             by = c("year", "stabbr")) %>%
  inner_join(cg %>% 
               select(year, capgains=cbo_jul2020) %>%
               mutate(cglagged=capgains[match(year -1, year)]),
             by="year") %>%
  pivot_longer(-c(year, stabbr)) %>%
  group_by(stabbr, name) %>%
  mutate(pch=value / value[match(year - 1, year)] * 100 - 100) %>%
  ungroup


```




```{r}
# construct dataframe that will hold forecasts
osrvars <- c("iit", "gst", "cit", "othertax", "nontax") # just the elements, not totals

# create a base data frame to hold forecasts - joins will grab only the states of interest
base <- scenarios %>%
  inner_join(stgdp_uswide %>% 
               rename(econfc=forecast),
             by = c("stabbr", "econfc")) %>%
  # create an indexed value we can use to get stgdp growth from any year to another year
  group_by(scenario, econfc, equityfc, govtfc) %>%
  mutate(istgdp=stgdp / stgdp[year==1965]) %>%
  ungroup %>%
  left_join(capgains %>% 
              rename(econfc=gdpname, equityfc=eqname),
            by = c("econfc", "equityfc", "year")) %>%
  full_join(tibble(revtype=osrvars), by=character()) %>%  # create a record for each revtype!
  left_join(osr %>% 
              filter(name %in% osrvars) %>% 
              rename(revtype=name, rev=value, rev_pch=pch),
            by = c("stabbr", "year", "revtype")) %>%
  select(stabbr, scenario, year, 
         stgdp, stgdp_pch, 
         usgdp, usgdp_pch, istgdp,
         capgains, capgains_lag, capgains_pch,
         revtype, rev, rev_pch,
         econfc, equityfc, govtfc)
base
summary(base)
count(base, stabbr)
count(base, stabbr, scenario, econfc, equityfc, govtfc)
count(base, revtype)

```


```{r assume2021}
# use state gdp to fill in nontax growth rates
assume2021 <- assume %>%
  filter(year==2021) %>%
  right_join(scenarios, by = c("stabbr", "govtfc")) %>%  # note the right join
  left_join(stusgdp %>% 
              filter(year==2021) %>%
              select(stabbr, year, econfc=forecast, gdppch=pch), 
            by = c("stabbr", "year", "econfc")) %>%
  mutate(nontax=ifelse(is.na(nontax), gdppch, nontax)) %>%
  select(-gdppch)
assume2021

assume2021_long <-
  assume2021 %>%
  pivot_longer(cols=c(iit, cit, gst, othertax, nontax), names_to = "revtype", values_to = "pch2021")

```



```{r fc_2021}
# fill in 2021 values using assume2021 growth rates

fcast <- base %>%
  left_join(assume2021_long %>%
              select(-scenario),
            by=c("year", "stabbr", "revtype", "govtfc", "econfc", "equityfc")) %>%
  group_by(stabbr, scenario, econfc, equityfc, govtfc, revtype) %>%
  mutate(rev=ifelse(year==2021 & is.na(rev),
                    rev[year==2020] * (1 + pch2021 / 100),
                    rev),
         rev_pch=rev / rev[match(year - 1, year)] * 100 - 100) %>%
  select(-pch2021) %>%
  ungroup
# glimpse(fcast)
# fcast %>% filter(year %in% 2020:2021, stabbr=="MA")


```


```{r}
# create a tsibble for model-based forecasts
fcast_ts <- fcast %>%
  select(-rev_pch) %>%
  filter(year >= 1996) %>% # we don't have capgains_lag before here
  mutate(stgdp_b=stgdp / 1e3) %>%  # put stgdp in billions to reduce numerical issues
  as_tsibble(index=year, key = c(stabbr, econfc, equityfc, govtfc, revtype))

```


```{r fc_fitsave_recovery_model, include=FALSE}
# estimate two simple regression with ARIMA errors models
#   gdp and capital gains, which we will use only for the income tax
#   gdp only, which we will use for other tax sources
a <- proc.time()
revmods <- fcast_ts %>%
  filter(year <= 2021) %>%
  model(modcg = ARIMA(rev ~ stgdp_b + capgains_lag),
        modbase = ARIMA(rev ~ stgdp_b))
b <- proc.time()
b - a  # takes about 16 seconds to estimate, so save
saveRDS(revmods, here::here("data", "revmods.rds"))

```

To illustrate the kinds of models estimated, we show the fit statistics for the income tax model for Massachusetts. The model includes state GDP (stgdp_b), lagged capital gains (capgains_lag), and two autoregressive terms to adjust for autocorrelation.


```{r describe_revmods}
revmods <- readRDS(here::here("data", "revmods.rds"))

revmods %>%
  filter(stabbr=="MA", revtype=="iit", econfc=="cbo_feb2021", govtfc=="gov_oct2020") %>%
  select(-modbase) %>%
  report()
  
# show the fit
# mod %>% 
#   augment() %>%
#   select(year, cg, fit=.fitted) %>%
#   pivot_longer(-year) %>%
#   ggplot(aes(year, value, colour=name)) +
#   geom_line() +
#   geom_point() +
#   ggtitle("Model fit for capital gains") +
#   theme_bw()

```



```{r fc_forecast_recovery_models}
# forecast taxes in the recovery period using models
# create fcast_mod with recovery years only

revmods <- readRDS(here::here("data", "revmods.rds"))
# forecast
recovery_modfc <- revmods %>%
  forecast(new_data=fcast_ts %>% filter(year %in% model_forecast_years))

fcast_mod <- recovery_modfc %>%
  as_tibble() %>%
  mutate(iitmod = (revtype=="iit") & (.model=="modcg"),
         othermod=(revtype != "iit") & (.model=="modbase")) %>%
  filter(iitmod | othermod) %>%
  select(stabbr, year, scenario, econfc, equityfc, govtfc, revtype, fcmod=.mean)
summary(fcast_mod)  # only includes the recovery years

fcast_mod %>% 
  filter(stabbr=="NJ", year %in% 2019:2025, scenario %in% c("baseline", "downside"))


```



```{r fc_forecast_recovery_gapclose}
# forecast using the gap-closing method
# create fcast_gap with recovery years only
targets

fcast_gap <- fcast %>%
  group_by(stabbr, scenario, revtype) %>%
  mutate(revgdp=rev / stgdp * 100,
         troughpct=revgdp[year==trough_year]) %>%
  filter(year %in% gdpgap_recovery_years) %>%
  left_join(targets %>% 
              filter(ma_years==1, year==2019) %>%
              select(stabbr, revtype=name, longrun_targetpct),
            by=c("stabbr", "revtype")) %>%
  mutate(troughgap=troughpct - longrun_targetpct,
         year_targetpct=troughpct - troughgap * (year - trough_year) / length(gdpgap_recovery_years),
         fcgap=stgdp * year_targetpct / 100) %>%
  ungroup
summary(fcast_gap)
glimpse(fcast_gap)

fcast_gap %>%
  filter(stabbr=="NJ", year <=2026) %>%
  select(stabbr, scenario, year, revtype, stgdp, rev, revgdp, 
         troughpct, longrun_targetpct, troughgap, year_targetpct, fcgap) %>%
  arrange(stabbr, scenario, revtype, year)

fcast_gap %>%
  filter(stabbr=="NJ", year <=2026) %>%
  select(stabbr, scenario, year, revtype, stgdp, troughpct, longrun_targetpct,
         troughgap, year_targetpct, fcgap) %>%
  arrange(stabbr, scenario, revtype, year)
  


```


```{r fc_longterm}
summary(fcast_mod)
summary(fcast_gap)

mainvars <- c("stabbr", "year", "econfc", "equityfc", "govtfc", "revtype")

fcast2 <- fcast %>%
  left_join(fcast_mod %>% select(all_of(c(mainvars, "fcmod"))),
            by = c("stabbr", "year", "revtype", "econfc", "equityfc", "govtfc")) %>%
  left_join(fcast_gap %>% select(all_of(c(mainvars, "fcgap"))),
            by = c("stabbr", "year", "revtype", "econfc", "equityfc", "govtfc")) %>%
  # fill in history
  mutate(across(c(fcmod, fcgap), ~ ifelse(year <= 2021, rev, .))) %>%
  # 2022+ use pch for nontax, all years
  group_by(stabbr, econfc, equityfc, govtfc, revtype) %>%
  # forecast nontax, all years, by assuming it grows as state gdp grows
  mutate(across(c(fcmod, fcgap), 
                ~ ifelse(year > 2021 & revtype=="nontax",
                         rev[year==2021] * istgdp / istgdp[year==2021],
                         .)),
         # use separate periods for fcmod and fcgap
         fcmod = ifelse(year > max(model_forecast_years) & revtype!="nontax",
                        fcmod[year==max(model_forecast_years)] * istgdp / istgdp[year==max(model_forecast_years)],
                        fcmod),
         fcgap = ifelse(year > max(gdpgap_recovery_years) & revtype!="nontax",
                        fcgap[year==max(gdpgap_recovery_years)] * 
                          istgdp / istgdp[year==max(gdpgap_recovery_years)],
                        fcgap)) %>%
  ungroup

# fcast2 %>% filter(stabbr=="NJ", year %in% 2020:2025) %>% select(year, rev, fcmod, fcgap)
fcast2 %>% 
  filter(stabbr=="NJ", year %in% 2020:2026, scenario=="baseline", revtype=="iit") %>% 
  select(year, revtype, stgdp_pch, istgdp, rev, fcmod, fcgap)
  
# long
names(fcast2)
fcast_long <- fcast2 %>%
  select(stabbr, year, scenario, 
         econfc, equityfc, govtfc,
         usgdp, usgdp_pch,
         stgdp, stgdp_pch,
         capgains, capgains_lag,
         capgains_pch,
         revtype, model=fcmod, gapclose=fcgap) %>%
  pivot_longer(c(model, gapclose), names_to = "method", values_to = "rev")


```


```{r fc_wide}
fcast_wide <- fcast_long %>%
  pivot_wider(names_from = revtype, values_from = rev) %>%
  mutate(tottax=naz(iit) + naz(gst) + naz(cit) + naz(othertax),
         osr=tottax + naz(nontax))
  
# fcast_wide %>%
#   filter(stabbr=="NJ", method=="model") %>%
#   filter(year >= 2015) %>%
#   ggplot(aes(year, osr, colour=scenario)) +
#   geom_line() +
#   geom_point() +
#   ggtitle("osr")
# 
# fcast_wide %>%
#   filter(stabbr=="NJ", method=="model") %>%
#   filter(year >= 2010) %>%
#   mutate(pct=osr / stgdp * 100) %>%
#   ggplot(aes(year, pct, colour=scenario)) +
#   geom_line() +
#   geom_point() +
#   ggtitle("pct")
# 
# fcast_wide %>%
#   filter(stabbr=="NJ", scenario=="precovid") %>%
#   filter(year >= 2010) %>%
#   mutate(pct=osr / stgdp * 100) %>%
#   ggplot(aes(year, pct, colour=method)) +
#   geom_line() +
#   geom_point() +
#   ggtitle("pct")
# 
# 
# fcast_wide %>%
#   pivot_longer(cols=c(iit, gst, cit, othertax, tottax, nontax, osr),
#                names_to = "revtype",
#                values_to = "rev") %>%
#   mutate(pctgdp=rev * 1000 / stgdp * 100) %>%
#   filter(stabbr=="NJ", year >= 2015, scenario=="baseline") %>%
#   ggplot(aes(year, pctgdp, colour=revtype)) +
#   geom_line() +
#   geom_point() +
#   ggtitle("pctgdp")

```




```{r state_gdp_functions}
# functions to be applied to each state
gdp_diff <- function(st){
    # compare national and state gdp growth for a state
  df <- stusgdp %>%
    filter(stabbr %in% c("US", st),
           forecast=="moodys_dec2020",
           year >= 1965) %>%
    select(year, stabbr, pch) %>%
    pivot_wider(names_from = stabbr, values_from = pch) %>%
    mutate(diff=.[[st]] - US)
  df
}

gdp_comp_plot <- function(st){
  # compare national and state gdp growth for a state
  startyear <- 1990
  
  df <- gdp_diff(st)
  
  recuse <- recessions %>%
    filter(rec_year >= startyear)
  
  df %>%
    select(-diff) %>%
    filter(year >= startyear, year <= 2030) %>%
    pivot_longer(-year, names_to = "stabbr") %>%
    ggplot(aes(year, value, colour=stabbr)) +
      geom_line(size=1) +
      geom_point() +
      scale_colour_manual(values=c("blue", "darkgreen")) +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 2020.5, linetype="dotted", size=1) +
      # geom_hline(aes(yintercept=avg$avg), colour="red", size=0.4) +
      scale_y_continuous(name="% change", breaks=seq(-20, 20, 1)) +
      scale_x_continuous(name="State fiscal year", breaks=seq(1980, 2050, 5)) +
      annotate("rect",
         fill = "lightgrey",
         alpha = 0.5, # larger alpha is darker rectangle
         xmin = recuse$peak_decimal, xmax = recuse$trough_decimal,
         ymin = -Inf, ymax = Inf) +
      ggtitle(paste0("Nominal GDP growth for ", st, " and the nation, Moody's December 2020 forecast"),
              subtitle=paste0("Dotted vertical line marks break between history and forecast")) +
      theme_bw()
}


gdp_diff_plot <- function(st){
  # compare national and state gdp growth for a state
  df <- gdp_diff(st)
  
  startyear <- 1990
  recuse <- recessions %>%
    filter(rec_year >= startyear)
  
  avgstart <- 2000
  avgend <- 2020
  avg <- df %>%
    filter(year %in% avgstart:avgend) %>%
    summarise(avg=mean(diff))
  
  df %>%
    filter(year >= startyear, year <= 2030) %>%
    ggplot(aes(year, diff)) +
      geom_line(colour="blue", size=1) +
      geom_point(colour="blue") +
      geom_hline(yintercept = 0) +
      geom_vline(xintercept = 2020.5, linetype="dotted", size=1) +
      geom_hline(aes(yintercept=avg$avg), colour="red", size=0.4) +
      scale_y_continuous(name="State % change minus US % change", breaks=seq(-10, 10, 1)) +
      scale_x_continuous(name="State fiscal year", breaks=seq(1980, 2050, 5)) +
      annotate("rect",
         fill = "lightgrey",
         alpha = 0.5, # larger alpha is darker rectangle
         xmin = recuse$peak_decimal, xmax = recuse$trough_decimal,
         ymin = -Inf, ymax = Inf) +    
      ggtitle(paste0("Nominal GDP growth: ", st, " minus nation, Moody's December 2020 forecast"),
              subtitle=paste0("Red horizontal line is average over ", avgstart, "-", avgend,
                              ". Dotted vertical line marks break between history and forecast.")) +
      theme_bw()
  # df
}


```


```{r revenue_functions}

gdp_osr_plot <- function(st){
  recuse <- recessions %>%
    filter(rec_year > 1978)

  p <- gdposr_pch %>%
    filter(stabbr==st, year >=1978) %>%
    ggplot(aes(year, pch, colour=name)) +
      geom_line(size=1) +
      geom_point() +
      scale_colour_manual(values=c("blue", "darkgreen")) +
      geom_hline(yintercept = 0) +
      annotate("rect",
             fill = "lightgrey",
             alpha = 0.5, # larger alpha is darker rectangle
             xmin = recuse$peak_decimal, xmax = recuse$trough_decimal,
             ymin = -Inf, ymax = Inf) +
      scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
      scale_y_continuous(name="% change vs. year ago", breaks=seq(-50, 50, 2)) +
      theme_bw() +
      theme(legend.title = element_blank())
  p
}

iit_gdpcg_plot <- function(st){
  stname <- stname(st)
  
  recuse <- recessions %>%
    filter(rec_year > 1995)
    
  p <- iit_gdpcg %>%
    filter(stabbr==st, year >= 1996, name != "capgains") %>%
    ggplot(aes(year, pch, colour=name)) +
      geom_line(size=1) +
      geom_point() +
      scale_colour_manual(values=c("blue", "darkgreen", "red")) +
      geom_hline(yintercept = 0) +
      annotate("rect",
             fill = "lightgrey",
             alpha = 0.5, # larger alpha is darker rectangle
             xmin = recuse$peak_decimal, xmax = recuse$trough_decimal,
             ymin = -Inf, ymax = Inf) +
      scale_x_continuous(name=NULL, breaks=seq(1960, 2030, 5)) +
      scale_y_continuous(name="% change vs. year ago", breaks=seq(-100, 100, 5)) +
      ggtitle(paste0(stname, ": % change in state income tax, state GDP, and lagged national capital gains")) +
      theme_bw() +
      theme(legend.title = element_blank())
  p
}


```


```{r forecast_reporting_functions}
# fcast_wide %>%
#   pivot_longer(cols=c(iit, gst, cit, othertax, tottax, nontax, osr),
#                names_to = "revtype",
#                values_to = "rev") %>%
#   mutate(pctgdp=rev * 1000 / stgdp * 100) %>%
#   filter(stabbr=="NJ", year >= 2015, scenario=="baseline") %>%
#   ggplot(aes(year, pctgdp, colour=revtype)) +
#   geom_line() +
#   geom_point() +
#   ggtitle("pctgdp")

osr_scen <- function(st, method){
  p <- fcast_wide %>%
    filter(stabbr==st, year >= 2010, method==!!method) %>%
  ggplot(aes(year, osr, colour=scenario)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(name="State fiscal year", breaks=seq(2000, 2040, 2)) +
  scale_y_continuous(name="Own-source revenue, $ millions", labels = scales::comma) +
  ggtitle(paste0("Scenarios using method: ", method)) +
  theme_bw()
  
  p
}

osr_method <- function(st, scenario, var){
  # if(st != "NJ" & scenario != "baseline"){
  #     print(paste0(scenario, " not yet available"))
  #     return()
  #   }  
  
  p <- fcast_wide %>%
    filter(stabbr==st, year >= 2010, scenario==!!scenario) %>%
    select(year, pvar=!!var, method) %>%
  ggplot(aes(year, pvar, colour=method)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(name="State fiscal year", breaks=seq(2000, 2040, 2)) +
  scale_y_continuous(name="$ millions", labels = scales::comma) +
  ggtitle(paste0("Methods for forecasting ", var, " using scenario: ", scenario)) +
  theme_bw()
  
  p
}


osr_method_pctgdp <- function(st, scenario, var){
  # if(st != "NJ" & scenario != "baseline"){
  #     print(paste0(scenario, " not yet available"))
  #     return()
  #   }  
  
  p <- fcast_wide %>%
    filter(stabbr==st, year >= 2010, scenario==!!scenario) %>%
    select(year, pvar=!!var, method, stgdp) %>%
  ggplot(aes(year, pvar / stgdp * 100, colour=method)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(name="State fiscal year", breaks=seq(2000, 2040, 2)) +
  scale_y_continuous(name="% of state GDP") +
  ggtitle(paste0("Methods for forecasting ", var, " using scenario: ", scenario)) +
  theme_bw()
  
  p
}


# st <- "NJ"
# scenario <- "baseline"

osr_bkdwn <- function(st, scenario, method){
  # if(st != "NJ" & scenario != "baseline"){
  #     print(paste0(scenario, " not yet available"))
  #     return()
  #   }
  
  stname <- stname(st)
  vars <- c("iit", "gst", "cit", "othertax", "tottax", "nontax", "osr")
  
  data <- fcast_wide %>%
    filter(stabbr==st, scenario==!!scenario, method==!!method, year >= 2018) %>%
    select(stabbr, year, all_of(vars))
  
  tab <- data %>%
  gt() %>%  
  tab_header(
      title = paste0(stname, " own-source revenue, $ millions"),
      subtitle = paste0("Scenario: ", scenario, "  --  Method: ", method)
    ) %>%
    cols_label(
      year=html("State fiscal year<br>ending in:")
    ) %>%
    fmt_currency(
      columns = vars,
      rows=1,
      decimals = 0,
      scale_by = 1,
      suffixing = FALSE
    ) %>%
    fmt_number(
      columns = vars,
      rows=2:nrow(data),
      decimals = 0,
      scale_by = 1,
      suffixing = FALSE
    ) %>%
    tab_style(
      style = list(
        cell_fill(color = "#f7f7f7")
      ),
      locations = cells_body(
        rows = seq(1, nrow(data), 2)
        )
    ) %>%
    tab_source_note("OSR categories use Census-definitions")
  
  tab
}


osr_pch <- function(st, scenario, method){
  # if(st != "NJ" & scenario != "baseline"){
  #     print(paste0(scenario, " not yet available"))
  #     return()
  #   }
  
  stname <- stname(st)
  vars <- c("iit", "gst", "cit", "othertax", "tottax", "nontax", "osr")
  
  data <- fcast_wide %>%
    filter(stabbr==st, scenario==!!scenario, method==!!method, year >= 2017) %>%
    select(stabbr, year, all_of(vars)) %>%
    mutate(across(all_of(vars), ~ . / .[match(year - 1, year)] * 100 - 100))

  tab <- data %>%
    gt() %>%  
    tab_header(
      title = paste0(stname, " own-source revenue, % change"),
      subtitle = paste0("Scenario: ", scenario, "  --  Method: ", method)
      ) %>%
      cols_label(
        year=html("State fiscal year<br>ending in:")
      ) %>%
      fmt_percent(
        columns = vars,
        # rows=1:nrow(data),
        decimals = 1,
        scale_values = FALSE
      ) %>%
      tab_style(
        style = list(
          cell_fill(color = "#f7f7f7")
        ),
        locations = cells_body(
          rows = seq(1, nrow(data), 2)
          )
      ) %>%
      tab_source_note("OSR categories use Census-definitions")
  
  tab
}


```



```{r report_function}
# define the functions we want to call, including titles
# workaround https://github.com/rstudio/rstudio/issues/1795#issuecomment-505779701
#   use plot.new(), dev.off()
#  cat('\n\n<!-- -->\n\n') may be needed if there are problems

state_report <- function(st){
  stname <- stname(st)
  
  cat("\n")
  cat("## ", stname, " \n")
  
  cat("\n")
  cat("### ", stname, " History \n")  #  {.tabset}

  # The economy
  cat("\n")
  cat("#### ", st, " GDP growth versus US", "\n")
  print(gdp_comp_plot(st))
  plot.new()
  dev.off()
  
  cat("\n")
  cat("#### ", st, " GDP growth minus United States GDP growth \n")
  print(gdp_diff_plot(st))
  plot.new()
  dev.off()
  cat('\n\n')
  
  cat("\n")
  cat("#### ", st, " State GDP growth and own-source revenue growth \n")
  print(gdp_osr_plot(st))
  plot.new()
  dev.off()
  cat('\n\n')
  
  cat("\n")
  cat("####  ", st, " State income tax, state GDP growth, and lagged national capital gains\n")
  print(iit_gdpcg_plot(st))
  plot.new()
  dev.off()
  cat('\n\n')
  
    
  cat("\n")
  cat("### ", stname, " Forecasts \n")  # {.tabset}
  
  # OSR forecasts
  cat("\n")
  cat("#### ", st, " Own-source revenue scenarios\n")
  cat("##### ", st, " Method: Simple model\n")
  print(osr_scen(st, method="model"))
  plot.new()
  dev.off()
  cat("##### ", st, " Method: Close gap between revenue as % of state GDP, and 2019 value\n")
  print(osr_scen(st, method="gapclose"))
  plot.new()
  dev.off()
  
  cat("##### ", st, " Methods comparison, $ millions, baseline scenario\n")
  print(osr_method(st, scenario="baseline", var="osr"))
  plot.new()
  dev.off()
  print(osr_method(st, scenario="baseline", var="iit"))
  plot.new()
  dev.off()
  print(osr_method(st, scenario="baseline", var="gst"))
  plot.new()
  dev.off()
  print(osr_method(st, scenario="baseline", var="cit"))
  plot.new()
  dev.off()
  print(osr_method(st, scenario="baseline", var="othertax"))
  plot.new()
  dev.off()
  
  
  cat("##### ", st, " Methods comparison, $ millions, downside scenario\n")
  print(osr_method(st, scenario="downside", var="osr"))
  plot.new()
  dev.off()
  print(osr_method(st, scenario="downside", var="iit"))
  plot.new()
  dev.off()
  print(osr_method(st, scenario="downside", var="gst"))
  plot.new()
  dev.off()
  print(osr_method(st, scenario="downside", var="cit"))
  plot.new()
  dev.off()
  print(osr_method(st, scenario="downside", var="othertax"))
  plot.new()
  dev.off()  
  
  cat("##### ", st, " Methods comparison, % of state GDP, baseline scenario\n")
  print(osr_method_pctgdp(st, scenario="baseline", var="osr"))
  plot.new()
  dev.off()
  print(osr_method_pctgdp(st, scenario="baseline", var="iit"))
  plot.new()
  dev.off()
  print(osr_method_pctgdp(st, scenario="baseline", var="gst"))
  plot.new()
  dev.off()
  print(osr_method_pctgdp(st, scenario="baseline", var="cit"))
  plot.new()
  dev.off()
  print(osr_method_pctgdp(st, scenario="baseline", var="othertax"))
  plot.new()
  dev.off()
  
  
  cat("##### ", st, " Methods comparison, % of state GDP, downside scenario\n")
  print(osr_method_pctgdp(st, scenario="downside", var="osr"))
  plot.new()
  dev.off()
  print(osr_method_pctgdp(st, scenario="downside", var="iit"))
  plot.new()
  dev.off()
  print(osr_method_pctgdp(st, scenario="downside", var="gst"))
  plot.new()
  dev.off()
  print(osr_method_pctgdp(st, scenario="downside", var="cit"))
  plot.new()
  dev.off()
  print(osr_method_pctgdp(st, scenario="downside", var="othertax"))
  plot.new()
  dev.off()    

  
  cat('\n\n')
  
  # OSR forecast breakdown, alternative scenarios and methods
  cat("\n")
  cat("#### ", st, " Table: Own-source revenue breakdown - pre-Covid\n")
  print(osr_bkdwn(st, scenario="precovid", method="model"))
  plot.new()
  dev.off()
    print(osr_bkdwn(st, scenario="precovid", method="gapclose"))
  plot.new()
  dev.off()
  
  cat("\n")
  cat("#### ", st, " Table: Own-source revenue breakdown - Covid baseline\n")
  print(osr_bkdwn(st, scenario="baseline", method="model"))
  plot.new()
  dev.off()
  print(osr_bkdwn(st, scenario="baseline", method="gapclose"))
  plot.new()
  dev.off()
  
  cat("\n")
  cat("#### ", st, " Table: Own-source revenue breakdown - downside\n")
  print(osr_bkdwn(st, scenario="downside", method="model"))
  plot.new()
  dev.off()
  print(osr_bkdwn(st, scenario="downside", method="gapclose"))
  plot.new()
  dev.off()

  cat("\n")
  cat("#### ", st, " Table: Own-source revenue breakdown, % change - pre-Covid\n")
  print(osr_pch(st, scenario="precovid", method="model"))
  plot.new()
  dev.off()
  print(osr_pch(st, scenario="precovid", method="gapclose"))
  plot.new()
  dev.off()  
    
  cat("\n")
  cat("#### ", st, " Table: Own-source revenue breakdown, % change - Covid baseline\n")
  print(osr_pch(st, scenario="baseline", method="model"))
  plot.new()
  dev.off()
  print(osr_pch(st, scenario="baseline", method="gapclose"))
  plot.new()
  dev.off()
  
  cat("\n")
  cat("#### ", st, " Table: Own-source revenue breakdown, % change - downside\n")
  print(osr_pch(st, scenario="downside", method="model"))
  plot.new()
  dev.off()
  print(osr_pch(st, scenario="downside", method="gapclose"))
  plot.new()
  dev.off()
  
}

```


```{r run_reports, include=TRUE, results='asis', message=FALSE, warning=FALSE}
# fig.keep='all', fig.show='asis'

state_report("MA")
state_report("MN")
state_report("NJ")

```


## Comparison to Pew forecasts

```{r}
pewosr
fcast_wide

comp <- bind_rows(fcast_wide %>%
                    select(stabbr, year, scenario, method, osr) %>%
                    mutate(type="boyd"),
                  pewosr %>%
                    mutate(method="pew",
                           type="pew"))

f <- function(scen, st){
  comp %>%
    filter(str_detect(scenario, scen), stabbr==st, method %in% c("pew", "model")) %>%
    filter(year >= 2010, year <= 2030) %>%
    ggplot(aes(year, osr, colour=type)) +
    geom_line() +
    geom_point() +
    ggtitle("Comparison of Pew OSR forecast to Boyd model forecast",
            subtitle = paste0("For ", scen, " scenario, for: ", stname(st)))
}


fgap <- function(scen, st){
  comp %>%
    filter(str_detect(scenario, scen), stabbr==st, method %in% c("pew", "gapclose")) %>%
    filter(year >= 2010, year <= 2030) %>%
    ggplot(aes(year, osr, colour=type)) +
    geom_line() +
    geom_point() +
    ggtitle("Comparison of Pew OSR forecast to Boyd gap-closing forecast",
            subtitle = paste0("For ", scen, " scenario, for: ", stname(st)))
}


```


```{r fccomp, include=TRUE}
f("baseline", "MN")
f("downside", "MN")

f("baseline", "NJ")
f("downside", "NJ")

fgap("baseline", "MN")
fgap("downside", "MN")

fgap("baseline", "NJ")
fgap("downside", "NJ")

```


# Goals and methods

## Goals

Goal is to forecast state own-source revenue (osr) and components using a method that:

*   is compatible with a state's own short-term forecasts,
*   stabilizes in the longer term to grow at the rate of state gdp, and
*   can be applied to individual states with a minimal amount of digging for new information


Key forecast elements:

*   State own-source revenue and major components
    +   using Census definitions
    +   state fiscal years (periods ending in June)
    +   nominal
    +   $ millions
    +   2021 through 2030
    
*   National economic forecast - uses one of the relatively current forecasts provided by Pew:
    +   CBO July 2020 (baseline)
    +   Moody's December 2020, p50 (baseline) - the most-current full forecast we have
    +   Moody's July 2020, p50
    +   Moody's July 2020, p90 (S3 - severe stress scenario)
    +   (We also have available a Moody's December 2019 baseline forecast, not used)
    +   Federal Reserve alternative severe forecast
    
*   State economic forecast:

    +   available state GDP forecasts are Moody's baseline forecasts released in December 2019, July 2020, and December 2020, each of which also has an associated Moody's national forecast
    +   to get state GDP forecasts under different national scenarios, I assume the difference between a state's growth rate and the national growth rate is the same as it is in the most-recent Moody's forecast (December 2020)

*   Equity return forecasts - factor into capital gains forecasts, which in turn affect the income tax. There are two equity return forecasts (domestic equities)
    +   pew_baseline -- the Pew Covid baseline -- **Note that this is not consistent with the CBO July 2020 economic forecast**
    +   pew_fedalt
    
*   Capital gains forecasts

    +   From a simple model based upon national GDP, S&P 500, and capital gains history
    +   Forecast varies depending on which national GDP forecast is used and which equity forecast is used
  

Overall approach for OSR and components:

*   2019 and earlier - Census
*   2020 - tax growth rates from Census quarterly data; nontax based on state GDP estimate
*   2021 uses growth rates from a state governmental forecast, or inferred from those forecasts; nontax revenue assumed to grow at rate of state GDP
*   2022-2024 taxes are from simple models:
    +   income tax based upon state GDP, national capital gains, and its own history
    +   other tax categories based upon state GDP and the tax's own history
    +   nontax revenue assumed to grow at the rate of GDP
*   2025-2030 - all osr components are a constant % of GDP - their presumed long-term average (thus, all osr components grow at the same rate as state GDP)

Scenarios: Three factors determine a scenario

*   Which national economic forecast is used
*   Which equity forecast is used
*   Which governmental short-term forecast is used

Some combinations of these will not make sense (e.g., fedalt equity forecast might not make sense to combine with the CBO pre-covid economic forecast).

Currently I am using the Moody's December 2020 economic forecast as the best economic forecast. I know you prefer CBO July. We should discuss.

Here are the scenarios I am using (as of now):


  stabbr govtfc      econfc         equityfc      scenario

  NJ     ols_sep2020 moodys_dec2020 pew_covidbase baseline
  NJ     gov_aug2020 fedalt         pew_fedalt    downside
  NJ     gov_bmxleg  cbo_precovid   pew_covidbase precovid
  
  MA     bhi_oct2020 moodys_dec2020 pew_covidbase baseline


Selected issues and considerations:

*   Uses Census definitions of OSR and thus is different from what policymakers in NJ are familiar with
*   Assumes all revenue sources move from their 2021 values to longer term values, as % of GDP; historically, other tax revenue has tended to decline as % of GDP until policymakers explicitly raise it
*   Because the 2021 OSR estimate is based on a fixed governmental forecast, and is independent of the economic forecast, it may seem odd in relation to a severe-downturn economic forecast for FY 2021 - in particular, OSR as a percent of GDP is unusually high in 2021 when using the Moody's p90 forecast (because gdp is so low in that forecast)

## Data sources [details to come]

*   US GDP history: BEA NIPA tables, quarterly converted to state fiscal year (July-June)
*   State GDP history: BEA data, spliced NAICS and SIC; quarterly converted to state fiscal year, adjusted
*   US GDP forecasts:
    +   CBO pre-covid
